Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-25 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revisions on abstract and introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2898 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 008c34f7c84d6000f50baf27532300881343f061

In many approaches to model circular data it is assumed that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution is predicted by a linear combination of
covariates. \cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing another specific transformation
function which maps the values of the real line to the unit circle. 
Moreover, the presented methodology does not only allow for the mean direction to be modeled 
depending on covariates while a constant estimate is obtained for the concentration parameter 
but also for the vice versa version. For each of these two approaches a separate model is introduced, 
however, they are also combined in a joint model (referred to as ``mixed model'').
%Moreover, the presented methodology allows for both distribution parameters, the mean
%direction and the concentration parameter of the von Mises distribution, to be
%modeled depending on covariates, not only separately but also in a
%combined model (also referred to ``mixed model''). 
In the employed iterative optimization algorithm the two sets of regression coefficients used
to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
%To obtain proper initial values it is recommended to apply two separate models 
%where either one of the parameters is fixed while the other is predicted employing covariates 
%in a GLM-type model.
While the combined model but also the two separate models are built on well elaborated theory 
the employed methodology has shown to be very challenging in its application, mainly due to the 
complexity encountered in optimizing the log-likelihood function which is not globally concave. 
Therefore, highly informative starting values are crucial for the model to converge at all 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
present an alternative approach to regression modeling of circular data by employing a GLM in a 
Bayesian framework, however, only the mean direction is regressed on covariates.
All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
are built on additive regression models.
Generally these provide a powerful toolbox for modeling particularly smooth effects,
however, in the case of circular data a possibly highly non-linear link function
can induce a more complex behavior of the modeled parameters as the location
parameter shifts closely towards the ``origin point'' $0$, also depending on the 
direction of the shift.
Moreover, not all types effects of may be modeled adequately. Especially abrupt changes 
might not be detected properly owing to the enforced additive structure and selected model 
specification such as the form of linear combinations.
%% FIXME: clear explanation of difficulties close to 0

@@

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the fitted density (red line) and estimated
location parameter (red hand) this example shows that reasonably well fitting
probabilistic models can be established for circular data by specifying an
appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood. However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types
of data, including circular data.
Already existing generalized linear or additive approaches to circular regression 
models employ linear combinations of the covariates to connect them to one (or more) 
distribution parameters. 
%For the mean direction~$\mu$ \cite{Gould:1969} proposes the model 
%\begin{equation}
%\mu = \mu_0 + \sum_j \beta_j\cdot x_j(\textit{mod } 2\pi)
%\end{equation}
%with a circular intercept $\mu_0$ and $x$ being a vector of covariates
%with the corresponding vector of coefficients $\beta$.
By following this idea and including a link function $g$ many other models
have been developed in the general form 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} x).
\end{equation}
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an intervall of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. The concentration parameter

%% FIXME: information on already existing generalized linear and additive approaches
%% with formulas (\mu = g(X \beta) + \mu_0)   FIXME: add mu_0?
%% clear description of difficulties

% -------------------------------------------------------------------
% Circular regression trees and forests
% -------------------------------------------------------------------
\section{Circular regression trees and forests}
\label{sec:tree_forest}

%% FIXME: what are we aiming at? which problems do we want to avoid/solve?
%% mu and kappa as constants in subgroups, no initial values
%% Lund: what do they present (including formulas)
%% compare, what is different, how do we improve this approach

>>>>

In many approaches to model circular data it is assumed that the circular
variable of interest follows a circular distribution, in particular the von
Mises distribution which is also known as the ``circular normal distribution''
(see Section~\ref{sec:prob_circ} for more details). This is also the basis
which early regression models for circular data are built on as they regress
the parameter(s) of the von Mises distribution on covariates. One of the first
regression models with a circular response variable and linear covariates was
presented by \cite{Gould:1969} where the circular mean, i.e., the location
parameter of the von Mises distribution is predicted by a linear combination of
covariates. 
\cite{Johnson+Wehrly:1978} took this idea one step further by
plugging in a link function such that the values of the linear predictor are
transformed to a restricted interval of length $2\pi$. 
%This generalized linear
%model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
%subsequently by \cite{Fisher:1993} introducing another specific transformation
%function which maps the values of the real line to the unit circle. 
This generalized linear
model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and
subsequently by \cite{Fisher:1993} introducing independent GLMs for either the 
location or scale parameter with appropriate link functions while keeping the other 
parameter constant. Additionally, they introduce a combined hetereoscedastic version 
where the location and scale parameter are alternately reestimated conditional on 
respective sets of covariates until convergence.
%Moreover, the presented methodology does not only allow for the mean direction to be modeled 
%depending on covariates while a constant estimate is obtained for the concentration parameter 
%but also for the vice versa version. For each of these two approaches a separate model is introduced, 
%however, they are also combined in a joint model (referred to as ``mixed model'').
%%Moreover, the presented methodology allows for both distribution parameters, the mean
%%direction and the concentration parameter of the von Mises distribution, to be
%%modeled depending on covariates, not only separately but also in a
%%combined model (also referred to ``mixed model''). 
%In the employed iterative optimization algorithm the two sets of regression coefficients used
%to model $\mu$ and $\sigma$ respectively are updated in alternation until convergence.
%%To obtain proper initial values it is recommended to apply two separate models 
%%where either one of the parameters is fixed while the other is predicted employing covariates 
%%in a GLM-type model.
%While the combined model but also the two separate models are built on well elaborated theory 
While all of these models are built on well elaborated theory 
the employed methodology has shown to be very challenging in its application, mainly due to the 
complexity encountered in optimizing the log-likelihood function which is not globally concave. 
Therefore, highly informative starting values are crucial for the model to converge at all 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. 
In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} 
present an alternative approach to regression modeling of circular data by employing a GLM in a 
Bayesian framework, however, only the mean direction is regressed on covariates.
All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
are built on additive regression models.
Generally these provide a powerful toolbox for modeling particularly smooth effects,
however, in the case of circular data a possibly highly non-linear link function
can induce a more complex behavior of the modeled parameters as the location
parameter shifts closely towards the ``origin point'' $0$, also depending on the 
direction of the shift.
Moreover, not all types of effects may be modeled adequately. Especially abrupt changes 
might not be detected properly owing to the enforced additive structure and selected model 
specification such as the form of linear combinations.
%% FIXME: clear explanation of difficulties close to 0

@@

Figure~\ref{fig:densities} depicts a distributional model fitted to circular
data employing the von Mises distribution. While the left panel only serves as
an illustration of the corresponding ``linear'' description restricted to an
interval of length $2\pi$ but consequently not accounting for periodicity, the
right panel displays an appropriate representation of a circular quantity on
the unit circle. Looking at the fitted density (red line) and estimated
location parameter (red hand) this example shows that reasonably well fitting
probabilistic models can be established for circular data by specifying an
appropriate distribution family such as the von Mises distribution and
estimating its parameters via maximum likelihood. However, this procedure
considers only the response variable. Including covariates can improve any
model remarkably as they might provide valuable additional information.
Therefore, (distributional) regression models are of high interest for all types
of data, including circular data.

In already existing generalized linear or additive approaches to circular regression modeling 
usually the location parameter $\mu$ of is regressed on 
covariates in the general form 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} x).
\end{equation}
with a circular intercept $\mu_0$ and $x$ being a vector of covariates
with the corresponding vector of coefficients $\beta$ that are to be estimated.
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an interval of length $2\pi$ by setting $g(x) = 2\pi \cdot F(x)$ for a cumulative 
distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
The concentration parameter $\kappa$ is mostly estimated as a constant, however,
\cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on covariates
employing the exponential function as a link function while $\mu$ is assumed to be a fixed
value. But regardless of whether $\mu$ or $\kappa$ depend on covariates,
both variations have shown to be very complex in their optimization procedure and hence
challenging in their application. Consequently, these difficulties intensify even more in 
a combined model where the sets of regression 
coefficients for the two distribution parameters $\mu$ and $\kappa$ respectively are reestimated 
alternatingly in an iterative optimization algorithm. Hence, regardless of difficulties in practice, 
the presented theory provides a first approach to probabilistic regression modeling of circular data, 
however, the parameters are not regressed simultaneously on covariates as for example in a more general
case by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
Moreover, the additive structure of the discussed models might impose a smooth effect even if the 
true underlying effect is an abrupt shift.
One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
is to employ tree-based models as presented in the following Section.
%% FIXME: better description of difficulties?

% -------------------------------------------------------------------
% Circular regression trees and forests
% -------------------------------------------------------------------
\section{Circular regression trees and forests}
\label{sec:tree_forest}
%% FIXME: what are we aiming at? which problems do we want to avoid/solve?
%% mu and kappa as constants in subgroups, no initial values
%% Lund: what do they present (including formulas)
%% compare, what is different, how do we improve this approach
